Modern Statistical Methods --- Example Sheet 3

### 2020-12-01 inverse_problems/C5.tex
almost end of termgit add .!

>>>>

\begin{definition}
	In the formulation of Bayes' formula: 
	\begin{enumerate}
		\item $Z(y) \ceq g_{Y}(y)$ is called the \emph{model evidence} or \emph{marginal likelihood};
		\item $L(y\mid u) \ceq g_{Y\mid U =u}(y)$ is called the \emph{(data) likelihood};
		\item $\mu_0 \ceq \PP(U \in \cdot)$ is called the \emph{prior measure};
		\item $\mu_\Rm{post} \ceq \PP(U \in \cdot \mid Y = y)$ is called the \emph{posterior measure}. 
	\end{enumerate}
\end{definition}

If we assume that $\nu_U \ceq \mu_0$ (we can choose $\nu_U$ freely so long as $\mu_0$ has a probability density w.r.t.\ $\nu_0$), we obtain from Bayes' formula the formulation
\[
\dv{\mu_\Rm{post}}{\mu_0}(u) = \frac{L(y \mid u)}{Z(y)} \qquad(\mu_0\text{-a.s.}).
\] 
If $\mu_0$ has a density with respect to some other measure $\nu$ (say, Lebesgue measure), then we obtain
\[
\dv{\mu_\Rm{post}}{\nu} = \dv{\mu_\Rm{post}}{\mu_0} \dv{\mu_0}{\nu} = \frac{L(y \mid u)}{Z(y)} \dv{\mu_0}{\nu}. 
\]
In practice, this means that computing the density of the posterior measure comes down to multiplying the likelihood with the prior measure and normalising accordingly.

### 2020-12-01 inverse_problems/inverse_ES3_handin.tex
almost end of termgit add .!

\begin{question}
	Let $(\Omega, \FF)$ be a measurable space and $\mu, \nu, \rho$ be $\sigma$-finite measures on $(\Omega, \FF)$. Prove the following statements.
	\begin{enumerate}[(a)]
		\item Let $\nu \ll \mu$ and $a \geq 0$. Then, $a \cdot \nu \ll \mu$ and $\dv{a\cdot \nu}{\mu} = a \dv{\nu}{\mu}$ ($\mu$-a.e.). 
		\item Let $\nu \ll \mu$ and $\rho \ll \mu$. Then, $\nu + \rho \ll \mu$ and $\dv{\nu + \rho}{\mu} = \dv{\nu}{\mu} + \dv{\rho}{\mu}$ ($\mu$-a.e.). 
		\item Let $\rho \ll \nu$ and $\nu \ll \mu$. Then, $\rho \ll \mu$ and $\dv{\rho}{\mu} = \dv{\rho}{\nu} \cdot \dv{\nu}{\mu}$ ($\mu$-a.e.). 
	\end{enumerate}
\end{question}

\begin{proof}
	\begin{enumerate}[(a)]
		\item We have, for $F \in \FF$, that $\mu(F) = 0 \implies \nu(F) = 0 \implies (a\cdot \nu)(F) = 0$. Furthermore, we have
		\[
		(a \cdot \nu)(F) = a \cdot (\nu(F)) = a \cdot \int_F \dv{\nu}{\mu} \dd{\mu} = \int_F \qty(a \dv{\nu}{\mu}) \dd{\mu}, 
		\]
		which proves the Radon-Nikodym derivative of $a\cdot\nu$ is $a \dv{\nu}{\mu}$.

		\item We have, for $F \in \FF$, that $\mu(F) = 0 \implies \nu(F) = 0 \text{ and } \rho(F) = 0$ and therefore also $(\nu + \rho)(F) =0$. Furthermore, we have
		\[
		(\nu + \rho)(F) = \nu(F) + \rho(F) = \int_F\dv{\nu}{\mu} \dd{\mu} + \int_F \dv{\rho}{\mu} \dd{\mu} = \int_F \qty(\dv{\nu}{\mu} + \dv{\rho}{\mu}) \dd{\mu}, 
		\]
		which proves the Radon-Nikodym derivative of $\nu + \rho$ is $\dv{\nu}{\mu} + \dv{\rho}{\mu}$.

		\item We have, for $F \in \FF$, that $\mu(F) = 0\implies \nu(F) = 0 \implies \rho(F) = 0$. Furthermore, we have
		\[
		\rho(F) = \int_F \dv{\rho}{\nu} \dd{\nu} \overset\star= \int_F \dv{\rho}{\nu} \dv{\nu}{\mu} \dd{\mu},
		\]
		which proves the Radon-Nikodym derivative of $\rho$ is $\dv{\rho}{\nu} \dv{\nu}{\mu}$. Here, $\star$ follows from the fact that if a measure $\nu$ has $\mu$-density $g$, then $\int_\XX f \dd{\nu} = \int_\XX fg \dd{\mu}$ for all $f \geq 0$ --- this can be proved first for simple functions, and then extended to nonnegative integrable functions. 
	\end{enumerate}
\end{proof}

\begin{question}
	Let $(\Omega, \FF, \PP)$ be a probability space, and $U, U' \colon (\Omega, \FF) \to (\RR, \BB\RR)$, $Y \colon (\Omega, \FF) \to (\YY, \BB\YY)$ random variables. Moreover, let $U$ and $U'$ be integrable. Prove the following statements:
	\begin{enumerate}[(a)]
		\item Let $c \in \RR$ and $\PP(U = c) = 1$. Then, $\EE[U \mid Y = y] = c$  ($\PP(Y \in \cdot)$-a.s.). 
		\item Let $c \in \RR$. Then $\EE[cU \mid Y = y] = c \EE[U \mid Y = y]$ ($\PP(Y \in \cdot)$-a.s.).
		\item $\EE[U + U' \mid Y = y] = \EE[U \mid Y = y] + \EE[U' \mid Y = y]$ ($\PP(Y \in \cdot)$-a.s.). 
	\end{enumerate}
\end{question}

\begin{proof}
	Let $F \in \BB\YY$.

	\begin{enumerate}[(a)]
		\item We have for all $F$
		\[
		%\int_F \EE[U \mid Y = y] \PP(Y \in \dd{y}) = 
		\int_\qty{Y \in F} U \dd{\PP} = \int_\qty{Y \in F} c \dd{\PP} = c \PP(Y \in F) = \int_F c \PP(Y \in \dd{y}).
		\]
		\item We have for all $F$
		\[
		\int_\qty{Y \in F} cU \dd{\PP} = c \int_\qty{Y \in F} U \dd{\PP} = c \int_F \EE[U \mid Y = y] \PP(Y \in \dd{y}) = \int_F c\EE[U \mid Y = y] \PP(Y \in \dd{y}).
		\]

		\item We have for all $F$
		\begin{align*}
		\int_\qty{Y \in F} \qty(U + U') \dd{\PP} &= \int_\qty{Y \in F} U \dd{\PP} + \int_\qty{Y \in F} U' \dd{\PP} \\
		&= \int_F \EE[U \mid Y = y] \PP(Y \in \dd{y}) + \int_F \EE[U' \mid Y = y] \PP(Y \in \dd{y}) \\
		&= \int_F \qty( \EE[U \mid Y = y] + \EE[U' \mid Y = y]) \PP(Y \in \dd{y}).
		\end{align*}
	\end{enumerate}
\end{proof}

@@

\begin{question}
	Let $\mu$ be a probability measure on $(\RR, \BB\RR)$ with cumulative distribution function (cdf) $F \colon \RR \to [0, 1] \colon x \mapsto \mu((-\infty, x])$, for $x \in \RR$. 
	\begin{enumerate}[(i)]
		\item Let $Q \colon (0, 1) \to \RR$ be the quantile function of $\mu$, i.e., $Q(y) \ceq \inf\qty{x \in \RR \mid F(x) \geq y}$. Moreover, let $U \sim \Rm{Unif}(0, 1) \ceq \lambda_1(\cdot \cap (0, 1))$ be a uniformly distributed random variable on the interval $(0, 1)$. Show that $\PP(Q(U) \in \cdot) = \mu$. (Hint: you may use the fact that probability measures on $(\RR, \BB\RR)$ are uniquely determined by their CDF). 
		\item Derive the quantile function for the exponential distribution, i.e., the distribution with cdf $F(x) \ceq 1 - \exp(-\lambda x)$ for some $\lambda > 0$.

		\item Use the idea from (i) and your quantile function from (ii) to generate independent samples (i.e., realisations of random variables) with $\lambda = 1$. 
		\begin{enumerate}[(a)]
			\item Plot the cdf of the exponential distribution along with the empirical cdf of 
			\[
			M \in \qty{10, 100, 1000, 10000}
			\]
			of your samples. (Hint: you can use the \verb|ecdf| command to obtain a representation of your empirical cdf). What do you observe?
			\item Compute the sample mean $\bar X_M$ of your exponentially distributed samples $X_1, \dotsc, X_m$; for $M = \qty{2^n \mid n = 1, \dotsc, 20}$. What do you observe? 
		\end{enumerate}

		\item Repeat (iii)(b) using the quantile function $Q'(y) \ceq \tan(\pi(y - 0.5))$ ($y \in \RR$). What do you observe?

		\item Let $X_1, \dotsc, X_M \sim \mu$ be i.i.d., and assume that $\Var(X_1)$ is finite. Show that
		\[
		\EE\qty[\qty(\EE[X_1] - \bar X_M)^2] = \frac{\Var(X_1)}{M}. 
		\]
		\item Can you recover the rate from (v) in your experiments in (iii)? What could be the issue in (iv)? 
	\end{enumerate}

\end{question}

\begin{proof}
	\begin{enumerate}[(i)]
		\item Since probability measures on $(\RR, \BB\RR)$ are uniquely determined by their CDF, we must show that the CDF of $\mu$ agrees with the CDF of $\PP(Q(U) \in \cdot)$, i.e., for all $x \in \RR$
		\[
		F(x) = \mu((-\infty, x]) = \PP(Q(U) \in (-\infty, x]). 
		\]
		For this, note that
		\begin{align*}
			\PP(Q(U) \in (-\infty, x]) = \PP(Q(U) \leq x) \overset\star= \PP(U \leq F(x)) = F(x), 
		\end{align*}

	\item For the exponential distribution, the cdf $F$ is an invertible function between $(0, \infty)$ and $(0, 1)$, so the quantile function is just the inverse. We have
	\[	
	y = 1 - e^{-\lambda x} \iff e^{-\lambda x} = 1 - y \iff -\lambda x = \ln(1 - y) \iff x = -\frac{\ln(1 - y)}{\lambda}, 
	\]
	so the quantile function is given by $Q(y) = -\frac{\ln(1 - y)}{\lambda}$.

	\item

	\begin{enumerate}[(a)]
		\item From approximately $M = 1000$, the empirical CDF aligns almost exactly with the true cdf.
		\item Keeps getting closer to 1. 
	\end{enumerate}

\item The means are all over the place and don't seem to converge to anything.

\item We have 
\begin{align*}
	\EE\qty[\EE[X_1]^2 - 2\EE[X_1] \bar X_M + \bar{X_M}^2] &= \EE[X_1]^2 - 2\EE[X_1] \EE[\bar X_M] + \EE[\bar X_M^2] \\
	&= \EE[X_1]^2 - 2\EE[X_1]^2 + \frac1{M^2} \sum_{ij} \EE[X_i X_j] \\
	&= - \EE[X_1]^2 + \frac1M \EE[X_1^2] + \frac{M-1}{M} \EE[X_1]^2 \\
	&= \frac{\EE[X_1^2] - \EE[X_1]^2}{M} = \frac1M \Var(X_1).  
\end{align*}

\item In (iv), the issue is that the distribution does not have finite mean or variance. 
	\end{enumerate}

\end{proof}
\end{document}

>>>>

\setcounter{question}{2}

@@

\end{document}

### 2020-12-01 modern_statistical_methods/C2.tex
almost end of termgit add .!

With the previous proposition, we can write characterise solutions to the Lasso: namely we have
\[
\hat\beta_\lambda^L = \argmin_\beta \frac1{2n} \norm{Y- X\beta}_2^2 + \lambda \norm{\beta}_1 = \argmin_\beta Q(\lambda)
\]
if and only if $0 \in \pt Q(\hat \beta_\lambda^L)$. A simple computation gives, with $A = \qty{k \mid \beta_k \neq 0}$, 
\begin{align*}
\pt Q(\beta) &= \pt(\frac1{2n} \norm{Y- X\beta}_2^2) + \pt(\lambda \norm{\beta}_1) \\
&= \qty{-\frac1n X\T(Y - X\beta)} + \qty{\lambda \nu : \norm{\nu}_\infty \leq 1, \nu_A = \sign(\beta_A)}. 
\end{align*}

@@

\subsubsection{Variable selection}
We consider the noiseless model $Y = X\beta^0$, where no randomness is involved. Let $S = \qty{k \mid \beta_k \neq 0}$, $N = \qty{1, \dotsc, p} \setminus S$, and assume that $S = \qty{1, \dotsc, s}$ (so $\beta_1, \dotsc, \beta_s$ are nonzero and $\beta_{s+1}, \dotsc, \beta_p$ are zero).

>>>>

Using the KKT conditions, we can write characterise solutions to the Lasso: namely we have
\[
\hat\beta_\lambda^L = \argmin_\beta \frac1{2n} \norm{Y- X\beta}_2^2 + \lambda \norm{\beta}_1 = \argmin_\beta Q(\lambda)
\]
if and only if $0 \in \pt Q(\hat \beta_\lambda^L)$. A simple computation gives, with $A = \qty{k \mid \beta_k \neq 0}$, 
\begin{align*}
\pt Q(\beta) &= \pt(\frac1{2n} \norm{Y- X\beta}_2^2) + \pt(\lambda \norm{\beta}_1) \\
&= \qty{-\frac1n X\T(Y - X\beta)} + \qty{\lambda \nu : \norm{\nu}_\infty \leq 1, \nu_A = \sign(\beta_A)}. 
\end{align*}

@@

From this proposition, it follows that the vector $\hat\nu$ from the KKT conditions is also unique with $\hat\nu = \frac1{\lambda n} X\T (Y - X\hat\beta)$. 
\subsubsection{Variable selection}
We consider the noiseless model $Y = X\beta^0$, where no randomness is involved. Let $S = \qty{k \mid \beta_k \neq 0}$, $N = \qty{1, \dotsc, p} \setminus S$, and assume that $S = \qty{1, \dotsc, s}$ (so $\beta_1, \dotsc, \beta_s$ are nonzero and $\beta_{s+1}, \dotsc, \beta_p$ are zero).

### 2020-12-01 modern_statistical_methods/msm_ES3.tex
almost end of termgit add .!

By the KKT conditions we have $\norm{X\T\theta}_\infty = \frac1n\norm{X\T(Y - X\hat\beta)}_\infty \leq \lambda$, and plugging in $\theta^*$ in $G$ gives
\begin{align*}
	G(\theta^*) &= \frac1{2n} \qty(\norm{Y}_2^2 - \norm{X\hat\beta}_2^2) = \frac1{2n} \qty(\norm{Y - X\hat\beta}_2^2 + 2 Y\T X \hat\beta - 2 \norm{X\hat\beta}_2^2) \\
	&= \frac1{2n}\norm{Y - X\hat\beta}_2^2 + \hat\beta\T \qty{\frac1n X\T(Y - X\hat\beta)} \\
	&\overset\star= \frac1{2n} \norm{Y - X\hat\beta}_2^2 + \lambda \norm{\beta}_1,
\end{align*}
where $\star$ follows from the KKT conditions. This shows that $\theta^*$ maximises $G$ over the objective set.  \TODO Uniqueness of $\theta^*$?

@@

	It follows that the solution path is piecewise linear with knots whenever the sign of the Lasso solution changes. Since the Lasso solution is 0 for $\lambda$ large enough, there can only be finitely many sign changes, which concludes the proof. (\TODO elaborate on last step)
\end{proof}

>>>>

By the KKT conditions we have $\norm{X\T\theta}_\infty = \frac1n\norm{X\T(Y - X\hat\beta)}_\infty \leq \lambda$, and plugging in $\theta^*$ in $G$ gives
\begin{align*}
	G(\theta^*) &= \frac1{2n} \qty(\norm{Y}_2^2 - \norm{X\hat\beta}_2^2) = \frac1{2n} \qty(\norm{Y - X\hat\beta}_2^2 + 2 Y\T X \hat\beta - 2 \norm{X\hat\beta}_2^2) \\
	&= \frac1{2n}\norm{Y - X\hat\beta}_2^2 + \hat\beta\T \qty{\frac1n X\T(Y - X\hat\beta)} \\
	&\overset\star= \frac1{2n} \norm{Y - X\hat\beta}_2^2 + \lambda \norm{\beta}_1,
\end{align*}
where $\star$ follows from the KKT conditions. This shows that $\theta^*$ maximises $G$ over the objective set.  Finally, $\theta^*$ is the unique maximum since we are maximising a strictly concave function over a convex set.

@@

	It follows that the solution path is piecewise linear with knots whenever the sign of the Lasso solution changes. Since there are $3^p$ possible signs and alternating is not possible, there are finitely many knots. 
\end{proof}
